NumPy---Array Computation for Python

### 2020-01-06 body.tex
Remove additional sections

% BIDS -- UCB
% https://www.moore.org/grant-detail?grantId=GBMF5447
% $645,020 in 2016
% https://sloan.org/grant-detail/8222
% $659,359 in 2017
% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
% http://doi.org/10.5281/zenodo.3585761
% http://doi.org/10.5281/zenodo.3585767
In 2017, NumPy received its first large grants totaling 1.3M USD from the
Gordon \& Betty Moore and the Alfred P. Sloan foundations.
These two grants focus on addressing the technical debt accrued over the years and
on setting in place standards and architecture to encourage more sustainable development.
% CZI -- NumFOCUS/QuanSight
% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
% NumPy and OpenBLAS received $195,000 in 2019
% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
NumPy received a third grant for 195K USD from the Chan Zuckerberg
Initiative at the end of 2019.
This grant focuses on better serving NumPy's large number of beginning
to intermediate level users and on growing the community of NumPy
contributors.
Finally, since May 2019 the project receives around 10K USD annually from
Tidelift, which is typically used to fund documentation and website
improvements.

@@

%\begin{figure}
%    \centering
%    \includegraphics[width=0.9\linewidth]{scripts/PRs-using-CURRENT_MAINTAINERS.pdf}
%    \caption{Number of pull requests merged into the NumPy master branch for each
%        quarter since 2012. The total number of PRs is indicated with the
%        lower blue area showing the portion contributed by current or previous
%        maintainers.}\label{fig:prs-over-time}
%\end{figure}
%
%NumPy is currently maintained by a group of 23 contributors with commit rights
%to the NumPy code base. Out of these, 17 maintainers were active in
%2019, 4 of whom were paid to work on the project full-time.
%Additionally, there are a few long term developers who contributed and maintain
%specific parts of NumPy, but are not officially maintainers.
%At a release cycle of about every half year, the five recent releases in the years
%2018 and 2019 have averaged about 450~pull requests~(PRs) each,\footnote{
%    Note that before mid 2011, NumPy development did not happen on \url{github.com}.
%    All data provided here is based on the development which happened through GitHub
%    pull requests. In some cases contributions by maintainers may not be categorized as such.}
%% Since 1.14.0 (based on changelog): 381 + 438 + 490 + 531 + 402; last is preliminary
%with each release attracting more than a hundred new contributors.
%Figure~\ref{fig:prs-over-time} shows the number of PRs merged into the NumPy
%master branch.
%Although the number of PRs being merged fluctuates,
%the plot indicates an increased number of contributions over the past
%years.

@@

\section*{Array function protocol}

A vast number of projects are built on NumPy;
these projects are consumers of the NumPy API.
Over the last several years, a growing number of projects are providers of
a \emph{NumPy-like API} and array objects targeting audiences with specialized
needs beyond NumPy's capabilities.
For example, the NumPy API is implemented by several popular tensor computation
libraries including CuPy\footnote{\url{https://cupy.chainer.org/}},
JAX\footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}},
and Apache MXNet\footnote{\url{https://numpy.mxnet.io/}}. 
PyTorch\footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
and Tensorflow\footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}
provide tensor APIs with NumPy-inspired semantics.
It is also implemented in packages that support sparse arrays
such as \code{scipy.sparse} and \code{pydata.sparse}.
Another notable example is Dask, a library for parallel computing in
Python.  Dask adopts the NumPy API and therefore presents a familiar
interface to existing NumPy users, while adding powerful abilities to
parallelize and distribute tasks.

The multitude of specialized projects creates the difficulty that consumers
of these NumPy-like APIs write code specific to a single project and do not support
all of the above array providers.
This is a burden for users relying on the specialized array-like, since
a tool they need may not work for them.
It also creates new challenges for end-users who need to transition
from NumPy to a more specialized array.
The growing multitude of specialized projects with NumPy-like APIs threatened
to again fracture the scientific Python community.

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
An earlier step in this direction was the implementation of the
\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
for most mathematical functions.\cite{NEP13}
In 2019 this was expanded more generally with the inclusion of the
\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
These two protocols allows providers of array-like APIs to be interoperable
with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
For the users relying on specialized array projects it means that even though
much code is written specifically for NumPy arrays and uses the NumPy API as
\code{import numpy as np}, it can nevertheless work for them.
%For example, here is how a CuPy GPU array can be passed through NumPy for
%processing, with all operations being dispatched back to CuPy:
%
%\begin{lstlisting}
%import numpy as np
%import cupy as cp
%
%x_gpu = cp.array([1, 2, 3])
%y = np.sum(x_gpu)  # Returns a GPU array
%\end{lstlisting}

\section*{Random number generation}

The NumPy \code{random} module provides pseudorandom numbers from a wide range of
distributions. In legacy versions of NumPy, simulated random values are produced
by a \code{RandomState} object that: handles seeding and state initialization;
wraps the core pseudorandom number generator based on a Mersenne Twister
implementation\footnote{to be precise, the standard 32-bit version of MT19937};
interfaces with the underlying code that transforms random bits into
deviates from other distributions; and supplies a singleton instance exposed in
the root of the random module.

The \code{RandomState} object makes a compatibility guarantee so that a fixed
seed and sequence of function calls produce the same set of values. This
guarantee has slowed progress since improving the underlying code requires
extending the API with additional keyword arguments. This guarantee continues to
apply to \code{RandomState}.

NumPy 1.17 introduced a new API for generating random numbers that use a more
flexible structure that can be extended by libraries or end-users. The new API
is built using components that separate the steps required to generate random
variates. Pseudorandom bits are generated by a bit generator. These bits are
then transformed into variates from complex distributions by a generator.
Finally, seeding is handled by an object that produces sequences of high-quality
initial values.

Bit generators are simple classes that manage the state of an underlying
pseudorandom number generator. NumPy ships with four bit generators. The default
bit generator is a 64-bit implementation of the Permuted Congruential Generator
\cite{pcg64} (\code{PCG64}). The three other bit generators are a 64-bit version
of the Philox generator\cite{random123} (\code{Philox}), Chris Doty-Humphrey's
Small Fast Chaotic generator\cite{practrand} (\code{SFC64}), and the 32-bit
Mersenne Twister\cite{mt19937} (\code{MT19937}) which has been used in older
versions of NumPy.\footnote{The
\href{https://github.com/bashtage/randomgen}{randomgen project} supplies a wide
range of alternative bit generators such as a cryptographic counter-based
generators (\code{AESCtr}) and generators that expose hardware random number
generators (\code{RDRAND})\cite{randomgen}.} Bit generators provide
functions, exposed both in Python and C, for generating random integer
and floating point numbers.

The \code{Generator} consumes one of the bit generators and produces variates
from complicated distributions. Many improved methods for generating random
variates from common distributions were implemented, including the Ziggurat
method for normal, exponential and gamma variates\cite{ziggurat}, and Lemire's
method for bounded random integer generation\cite{lemire}. The \code{Generator}
is more similar to the legacy \code{RandomState}, and its API is substantially
the same. The key differences all relate to state management, which has been
delegated to the bit generator. The \code{Generator} does not make the same
stream guarantee as the \code{RandomState} object, and so variates may differ
across versions as improved generation algorithms are
introduced.\footnote{Despite the removal of the compatibility guarantee, simple
reproducibility across versions is encouraged, and minor changes that do not
produce meaningful performance gains or fix underlying bug are not generally
adopted.}

Finally, a \code{SeedSequence} is used to initialize a bit generator. The seed
sequence can be initialized with no arguments, in which case it reads entropy
from a system-dependent provider, or with a user-provided seed. The seed
sequence then transforms the initial set of entropy into a sequence of
high-quality pseudorandom integers, which can be used to initialize multiple bit
generators deterministically. A key design goal of a seed sequence was to be
splittable in the sense that a seed sequence can be used to produce child
sequences that are distinct from their parents or other ancestors. This
capability allows a seed sequence to be used in large distributed applications
where the number of workers required is not known. The sequences generated from
the same initial entropy and the same splits are fully deterministic to ensure
reproducibility.

The three components are combined to construct a complete random number
generator.
%
%\begin{lstlisting}
%from numpy.random import (
%    Generator,
%    PCG64,
%    SeedSequence,
%)
%
%seq = SeedSequence(1030424547444117993331016959)
%pcg = PCG64(seq)
%gen = Generator(pcg)
%\end{lstlisting}
%
This approach retains access to the seed sequence which can then be
used to spawn additional generators.
%
%\begin{lstlisting}
%children = seq.spawn(2)
%gen_0 = Generator(PCG64(children[0]))
%gen_1 = Generator(PCG64(children[1]))
%\end{lstlisting}
%
While this approach retains complete flexibility, the method
\code{np.random.default\_rng} can be used to instantiate a \code{Generator} when
reproducibility is not needed.

The final goal of the new API is to improve extensibility. \code{RandomState} is
a monolithic object that obscures all of the underlying state and functions. The
component architecture is one part of the extensibility improvements. The
underlying functions (written in C) which transform the output of a bit
generator to other distributions are available for use in CFFI. This allows the
same code to be run in both NumPy and dependent that can consume CFFI, e.g.,
Numba. Both the bit generators and the low-level functions can also be used in C
or Cython code.\footnote{As of 1.18.0, this scenario requires access to the
NumPy source. Alternative approaches that avoid this extra step are being
explored.}

@@

% 1st grant done, technical cleanup finished except dtypes
% 2nd grant starting, community growth and new user focus

Most of the technical work for the Moore and Sloan grant is now complete.
The main remaining item is an overhaul of the data type system, which
we are currently working on.
The meaning of each element within a NumPy array is described by its
datatype (dtype). NumPy arrays can hold all common numerical
datatypes, strings, datetimes, and generic Python objects, each of
which is identified by a dtype.
%We are busy overhauling the datatype
%system to make it behave consistently and to simplify the creation of
%custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
%of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
While numerical dtypes make the foundation for most users,
there are a multitude of use cases which cannot prosper due to current
limitations, such as physical units\cite{astropy,Goldbaum2018,pint},
% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
geometrical objects\cite{pygeos}, and automatic
differentiation\cite{pyadolc}.
%Projects addressing these exist, each struggles with current
%limitations.  % TODO: I may need citations, we could link github issues.
The proposed overhaul of the datatype system will make it behave consistently and
to simplify the creation of custom dtypes, both in C and in Python

%% While it is currently possible to define custom dtypes within NumPy, many of the
%% use cases above cannot be solved using this system.
%% This is evidenced also by code within NumPy which would require changes
%% in many places to add such dtypes, even when adding them within NumPy itself.
%% The new implementation will be simpler and more cleanly implemented in
%% this regard.
%% Custom dtypes and the basic dtypes included within NumPy will be implemented
%% and behave identically as much as possible.
%% At the same time redesigning the API will make NumPy dtypes more flexible to grow
%% to future needs should they arise.

%% As a further step, to make dtypes more accessible and allow a faster adoption of
%% custom dtypes within the community, we plan to create a Python API to
%% define such custom dtypes.

Having made progress on addressing technical debt accrued over two decades of
development and modernizing the code to handle new use cases, the project
is ready to grow its community of contributors and to better address the needs
of an ever-growing number of new users.  This is the main focus of the
forthcoming CZI grant.

[... CZI grant ...]

[... welcome new users ...]

% technical focus
% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays). 
% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.

% \subsubsection*{Future Development}

% Dtypes

% outline plan for the CZI grant and an invitation to new contributors

%\subsection*{Website}

>>>>

In 2018, we hired our first full-time developers ...

@@

@@

- array function protocol

- random number generation

Ongoing work

- dtype

@@
